Mathematical Aspects of the Capital Market

### 2017-08-25 phd/Chapterintroduction.tex
Fix a minor error.
Probably the final revision of the thesis.

\section{Regular Variation}\label{sec:intro_regvar}
The concept of regular variation is defined by the following scaling
property: if a function $f$ on $(0, \infty)$ satisfies
\[
\lim_{x \to \infty} {f(c x) \over f(x)} = c^\alpha
\quad
\forall c > 0,
\]
then we say $f$ is regularly varying with index $\alpha$.
$f$ can be written in the form $f(x) = {L(x) \over x^\alpha}$, where
$L(x)$ is a slowly varying function, i.e.
$\lim_{x \to \infty} {L(c x) \over L(x)} = 1$, $\forall c > 0$.
We call a random variable $X$ regularly varying with index
$\alpha \geq 0$ if it satisfies the tail balance condition in the
limit $x \to \infty$.
\[
\P(X > x) \sim p_+ {L(x) \over x^\alpha},\quad
\P(X < -x) \sim p_- {L(x) \over x^\alpha}, \quad
\text{for some } p_+, p_- \geq 0,\; p_+ + p_- = 1
\]

% Distribution functions, say $F(\cdot)$, whose survival function
% $\bar F(x) = 1 - F(x)$ satisfies the above scaling property, are
% observed in a variety of phenomena, 
When expanded to multiple dimensions, the scaling property of regular
variation is better described in terms of vague convergence to a
spectral measure $\mu_\alpha$: if a random vector $X$ satisfies
\[
{
  \P(X/|X| \in \cdot, |X| > c x)
  \over
  \P(|X| > x)
}
\overset{v}{\to} c^{-\alpha} \mu_\alpha(\cdot),
\quad
x \to \infty, \forall c > 0,
\]
then we say $X$ is regularly varying with index $\alpha$. Here
$\mu_\alpha$ is a probability measure on the unit sphere
\cite{buraczewski:damek:mikosch:2016}. It is called the spectral 
measure of $X$ and $\alpha$ is again the tail index.
If $X$ is regularly varying with index $\alpha$, then each component
and each linear combination of its components are regularly
varying with the same index $\alpha$. This follows from Feller
\cite{feller}, p. 278. Cf. also Jessen and Mikosch
\cite{JessenMikosch2006}, lemma 3.1, and Embrechts et. al.
\cite{embrechts:klueppelberg:mikosch:1997}, lemma 1.3.1.

@@

Random variables with regularly varying tails have some very nice
features: if $X_1$ and $X_2$ are both positive and regularly varying
with indices $\alpha_1$ and $\alpha_2$, respectively, then $a_1 X_1 + a_2 X_2$,
for $a_1, a_2 > 0$, is regularly varying with index
$\min\{\alpha_1, \alpha_2\}$ (cf. Mikosch and Jessen
\cite{JessenMikosch2006}). Moreover, if $X_1, X_2$ are iid,
$\P(a_1 X_1 + a_2 X_2 > u) \sim \P(a_1 X_1 > u) + \P(a_2 X_2 > u)$.

@@

In addition to Breiman's result, the following is also well-known,
assuming $X_1, X_2$ are positive independent random variables and
$X_1$ is regularly varying with index $\alpha$:
\begin{enumerate}
\item if $X_2$ is regularly varying with index $\alpha$ or
  $\lim_{x \to \infty} {P(X_2 > x) \over \P(X_1 > x)} = 0$, then $X_1
  X_2$ is regularly varying with index $\alpha$.
\item if $X_1, X_2$ are iid and $\E X_1^\alpha = \infty$, then
  ${\P(X_1 X_2 > x) \over \P(X_1 > x)} \to \infty$
\item if $X_1, X_2$ are iid and $\E X_1^\alpha < \infty$, then the
  only possible limit of ${\P(X_1 X_2 > x) \over \P(X_1 > x)}$ is
  $2 \E X_1^\alpha$.
\end{enumerate}
For an extensive summary of the regular variation properties of
functions of regularly varying random variables, see Mikosch and
Jessen \cite{JessenMikosch2006}.

@@

\section{Stochastic Recurrence Equation}
\label{sec:tyhyjyt}
One of the most important dynamical mechanisms that lead to regularly
varying random vectors is a stochastic recursion of the following form:
\begin{equation}
  \label{eq:rhjyu}
  X_t = A_t X_{t-1} + B_t, \quad t \in \integers
\end{equation}
where $X_t$ is a $d$-dimensional random vector, $A_t$ is a $d\times d$
random matrix and $B_t$ is a $d$-dimensional vector, random or
deterministic. The sequence $\{(A_t, B_t)\}_{t \in \integers}$ is
iid. The stationary solution to \eqref{eq:rhjyu} satisfies the fixed-point
equation $X \overset{d}{=} A X + B$, where $X$ and $(A, B)$ are
generic elements of the $\{X_t\}_{t \in \integers}$ and
$\{(A_t, B_t)\}_{t \in \integers}$ sequences.

@@

For discrete time models, Taylor \cite{taylor:1982} was the first to
propose a product process for modeling the martingale part $M_t$ of
$Y_t$:
\[
M_t - M_{t-1}= \sigma_t Z_t, \quad t \in \integers
\]
A convenient choice of $\sigma_t$ is
\begin{equation}
  \label{eq:rfht}
  \log\sigma_t = \sum_{l \in \integers} \psi_l \eta_{t-l}
  \quad
  t \in \integers,
\end{equation}
where $\{\psi_l\}_{l \in \integers}$ is a sequence of real numbers
with at least one non-zero element,
$\{\eta_t\}_{t \in \integers}$ is an iid sequence of random
variables with zero-mean and finite variance.
By Kolmogorov's 3-series theorem, the infinite series above
converges if and only if $\sum_{l \in \integers} \psi_l^2 < \infty$,
and in this case $\{\sigma_{t}\}$ is stationary.
In particular, if $\{\eta_i\}_{i \in \integers}$ is normally
distributed with zero mean, $\{\sigma_t\}_{t \in \integers}$ has
log-normal marginal distributions.

@@

There are a few advantages in using stochastic volatility models.
They are among the simplest models allowing for conditional
heteroscedasticity (cf. Andersen et al.
\cite{andersen:davis:kreiss:mikosch:2009}); nevertheless,
they greatly improve the accuracy of predicting future volatilities
by taking advantage of high frequency data. Specifically, for the
semi-martingale $\{Y_t\}$ we have
\begin{equation}
  \label{eq:jieg}
  [Y_{\delta}]_t
  =
  \sum_{j=1}^{\floor{t/\delta}}
  (Y_{j \delta} - Y_{(j-1)\delta})^2
  \overset{P}{\to} [Y]_t,
  \quad
  \delta \to 0.
\end{equation}
The object $[Y_\delta]_t$ is called the realized quadratic variation
in time series literature.
The convergence in probability follows directly from the definition of
quadratic variation $[Y]_t$:
\[
[Y]_t = \lim_{\Delta t \to 0} \sum_{i=1}^{n} (Y_{t_i} - Y_{t_{i-1}})^2
\]
where $0 = t_0 < t_1 < t_2 < \cdots < t_n = t$ and
$\Delta t = \sup_{1\leq i \leq n} (t_i - t_{i-1})$.
%% When $\E Z_t = 0$, $[Y]_t = [M]_t$ 
Furthermore, Jacod \cite{jacod:1994} and
Barndorff-Nielsen and Shephard \cite{barndorff:shephard:2002} proved a
central limit theorem:
\[
  {[Y_\delta]_t - [Y]_t \over \sqrt{2 \delta \int_0^t \sigma_s^2 ds}}
  \overset{d}{\to} N(0, 1),
  \quad \delta \to 0.
\]
If the process of expected return has continuous sample path and $M_t$
is a stochastic volatility process, it has been shown $[Y]_t = [M]_t$ and
$[Y_\delta]_t \overset{P}{\to} \int_0^t \sigma_s^2 ds$. Meanwhile,
It\^o's formula gives, for semi-martingale $Y_t$,
\begin{eqnarray}
  Y_t^2 &=& [Y]_t + 2 \int_0^t Y_s d Y_s \nonumber \\
  &=& [Y]_t + 2 \int_0^t Y_s dA_s + 2 \int_0^t Y_s dM_s, \nonumber \\
  \E Y_t^2 &=& \E [Y]_t + 2 \E \left( \int_0^t Y_s dA_s \right) \nonumber\\
  &\approx& \E [Y]_t, \quad \text{when $t$ is small}. \label{eq:hytre}
\end{eqnarray}
\eqref{eq:hytre} and \eqref{eq:jieg} show that $\E Y_t^2$, or in other
words, the forecast of future squared return, can be obtained as
$\E [Y_\delta]_t$, i.e. the forecast of future realized quadratic
variation.
% Since past realized quadratic variations are observable,
% forcast of future $[Y_\delta]_t$ can be obtained using e.g. an ARFIMA
% model.

@@

Conditional on $(Y_{t-1}, l_{t-1})$, the only source of randomness to
$Y_t$ comes from $Z_{t-1}$. Hence the shift of conditional probability
distribution shown in \eqref{eq:pgrtgpd} is equivalent to shifting the
distribution of $Z_{t-1}$. We have
\[
\P^\alpha (Z_{t-1} \in dz | Y_{t-1} = w) = |A(z^2) w|^\alpha {
  r_\alpha(A(z^2) \cdot w) \over r_\alpha(w)
} \P(Z_{t-1} \in dz).
\]
Note that $\{Z_t\}$ is an iid sequence in the original measure.
An expected value with respect to $\P^\alpha$ is related to its
counterpart with respect to $\P$ via
\begin{equation}
  \label{eq:pkjrkpo}
  \E [g(Y_t, l_t)] = \E^\alpha \left[
    g(Y_t, l_t) e^{-\alpha l_t} {r_\alpha(Y_{t-1}) \over r_\alpha(Y_t)}
    \right].
\end{equation}
Let us define $T_x = \min\{t \geq 1: |V_t| > x\}$. 
Once the first excursion of $|V_t|$ above $x$ has occurred, i.e.
$t > T_x$, we change the transition kernel back to the original and
continue the simulation until the process returns to a designated set
$\mathcal C = \{v: |v| \leq M\}$, where
the positive number $M$ is chosen in accordance with the function
$\Lambda$. We denote the successive times of $\{V_t\}$ returning to the
set $\mathcal C$ as $0 = K_0  < K_1 < K_2 < \cdots$. It can be shown that
$\{(V_{K_{m+1}}, \sum_{i=K_m}^{K_{m+1} - 1} \1{|V_i| > x})\}_{m \geq 0}$
is a positive Harris recurrent Markov chain for all $x \geq 0$.
Let $N_x = \sum_{i=0}^{K_1 - 1} \1{|V_i| > x}$. We show by the law of large
numbers for Markov chains
\[
\P(|V| > x) = \pi(\mathcal C) \E_\eta(N_x)
\]
where $\eta$ is $\pi$ restricted to the set $\mathcal C$, i.e.
$\forall S \subseteq \mathcal C, \eta(S) = \pi(S)/\pi(\mathcal C)$.
$\E_\eta$ means the expectation is taken only on condition
$V_0 \sim \eta$. Finally by \eqref{eq:pkjrkpo},
\[
\P(|V| > x) = \pi(\mathcal C) \E_\eta N_x = \E_\eta^{\alpha} \mathcal E_x
\]
where
\[
\mathcal E_x
=
\pi(\mathcal C)
N_x |A_{T_x} \cdots A_1 V_0|^{-\alpha}
{r_\alpha(Y_0) \over r_\alpha(Y_{T_x})}
\1{T_x < K_1}
\]
is our importance sampling estimator. $\E_\eta^{\mathcal D}$,
$\mathcal D$ for ``dual'', is to remind us that the expectation is
taken with respect to the shifted transition kernel as given by
\eqref{eq:pkjrkpo}, and then with respect to the original transition
kernel. In \S\ref{sec:efficiency} we show
\[
\limsup_{x \to \infty} {
  \E_\eta^{\mathcal D} \mathcal E_x^2
  \over
  [\P(|V| > x)]^2
} < \infty
\]
In plain words, the estimator $\mathcal E_x$ has bounded relative
error. See Asmussen and Glynn \cite{opac-b1123521}, \S 1 for
definition of {\em bounded relative error}.
The method presented in this paper is a multivariate generalization of
Collamore et al. \cite{collamore2014}. The reader is referred to it
for the treatment of the one-dimensional problem.
%% We propose an efficient importance sampling estimator for the rare
%%   event probability $\P(|V| > u)$ where $V$ is a vector following the
%%   stationary distribution of a GARCH($p, q$) process. Recall
%%   $V_t = A_t V_{t-1} + B$ is the matrix recurrence equation of a
%%   GARCH($p, q$) process. We emanate the process $\{V_t\}$ from a set
%%   $\mathcal C = \{V: |V| \leq M\}$ for some predefined positive
%%   constant $M$, and then we introduce a dual change of measure for the
%%   originally iid matrices $\{A_t\}$: for
%%   $t < T_u = \min\{i > 0: |V_i| > u\}$, we exponentially tilt the
%%   distribution of $A_t | X_{t-1}$, where $X_0 = V_0$ and
%%   $X_t = \prod_{i=1}^t X_0/|\prod_{i=1}^t X_0|$, so that
%%   $|A_t X_{t-1}|^\xi$ is more likely to take on large values and hence
%%   $|V_t|$ is more likely to exceed $u$. Once the exeedance has
%%   happened, we change the distribution of $\{A_t\}$ back to the
%%   original and continue the process until $V_t$ returns to the set
%%   $\mathcal C$. Along each simulated path emenating from $\mathcal C$
%%   and ending in $\mathcal C$ we compute $N_u = \sum_{i=1}^K \1{|V_i| >
%%     u}$, where $K = \min\{i > 0: |V_i| \leq M\}$.

@@

\subsection{Eigenvalues of the sample covariance matrix of a   
  stochastic volatility model}
In \S\ref{sec:SV:intro} we have introduced stochastic volatility
models as well as their connection to realized quadratic
variation and their improved predictive power derived thereby. But
there we have only discussed univariate models. In fact, the
generalization to multivariate models is rather straightforward. In
chapter \ref{ch:bernoulli} we adopt the following multivariate model:
\begin{eqnarray*}
  X_{i,t} &=& \sigma_{i,t} Z_{i,t} \quad
  1 \leq i \leq p,\;
  t \in \integers, \\
  \log \sigma_{i, t}
  &=&
  \sum_{k,l \in \integers} \psi_{k,l} \eta_{i-k, t-l}\quad
  1 \leq i \leq p,\;
  t \in \integers,
\end{eqnarray*}
where $\{Z_{i,t}\}$ and $\{\eta_{i,t}\}$ are iid fields of random
variables. They are independent of each other. The distribution of
$\eta$, a generic element of $\{\eta_{i,t}\}$, satisfies
$\P(e^\eta > x) \sim x^{-\alpha} L(x)$,
where $\alpha > 0$ and $L(x)$ is a slowly varying function. The
coefficients $\{\psi_{k,l}\}$ are real and satisfy
$\sum_{k, l \in \integers} |\psi_{k,l}| < \infty$.

Depending on the tails of $\{\sigma_{i,t}\}$ and $\{Z_{i,t}\}$,
two situations can arise. When $\{Z_{i,t}\}_{t \in \integers}$
is a regularly varying sequence with index $\alpha \in (0, 4)$
and dominates the tail, we show that each of the sequences
$\{X_{i,t}\}_{t \in \integers}, 1 \leq i \leq p$ and each of
$\{X_{i,t} X_{j,t}\}_{t \in \integers}, 1\leq i,j \leq p$
are regularly varying with index $\alpha$, assuming suitable
conditions on $\{\sigma_{i,t}\}$.

@@

We are interested in the model
\begin{eqnarray}
  X_{i,t} &=& \sum_{k,l \in \integers} h_{k,l} Z_{i-k, t-l},
  \quad i,t \in \integers \label{eq:tyhjuj}
\end{eqnarray}
where $\{Z_{i,t}\}_{i,t \in \integers}$ is a field of iid random
variables and $h_{k,l}$ is an array of real coefficients. We
assume $\{Z_{i,t}\}$ is regularly varying with index
$\alpha \in (0, 4)$ and
\[
\sum_{k,l \in \integers} |h_{k,l}|^\delta < \infty
\]
for some $\delta \in (0, \min\{\alpha/2, 1\})$. This conditions ensures
that the infinite series in \eqref{eq:tyhjuj} is almost surely
absolutely convergent. Since each $\{X_{i,t}\}_{t \in \integers}$
is a linear combination of the sequences $\{Z_{i,t}\}_{t \in \integers}$
that are regularly varying with index $\alpha$, each of
$\{X_{i,t}\}_{t \in \integers}$ is regularly varying with index $\alpha$.

@@

Corresponding to the time-lagged matrices $\mathbf X(s)$, we also introduce
the time-lagged coefficient matrices. Define
\[
\mathbf H(s) = \{h_{k, l+s}\}_{k, l \in \integers}
\]
and denote by $v_1(s) \geq v_2(s) \geq \cdots$ the singular values of
the matrix
$\mathbf H(0) \mathbf H(s) = \{\sum_{l \in \integers} h_{i,l} h_{j,l+s}\}_{i,j \in \integers}$.
We have established the following point process convergence result
upon appropriate conditions (see theorem \ref{cor:1} of chapter
\ref{ch:extremes}):
\[
\sum_{i=1}^p \varepsilon_{a_{np}^{-2}(\lambda_{(i)}(0), \ldots, \lambda_{(i)}(s))}
\overset{v}{\to}
\sum_{i=1}^\infty \sum_{j=1}^\infty
\varepsilon_{\Gamma_i^{-2/\alpha}(v_j(0), \ldots, v_j(s))}.
\]
where ${a_{np}}_{n,p \in \integers}$ is a sequence of positive real
numbers such that $\P(Z > a_{np}) = \ell(np) (np)^{-\alpha}$.
$\varepsilon_{x}$ is the Dirac measure with unit mass at $x$ and
$\Gamma_i = \sum_{k=1}^i E_k$, where $\{E_k\}_{1 \leq k \leq i}$ is
an iid sequence of Exp($1$) random variables. In particular, we see
$a_{np}^{-2}\lambda_{(1)}(0) \overset{d}{\to} E_1^{-2/\alpha} v_1(0)$, i.e.
asymptotically $a_{np}^{-2}\lambda_{(1)}(0)$ has a scaled
Fréchet($\alpha/2$) distribution.

>>>>

\section{Regular variation}\label{sec:intro_regvar}
The concept of regular variation is defined by the following scaling
property: if a function $f$ on $(0, \infty)$ satisfies
\[
\lim_{x \to \infty} {f(c x) \over f(x)} = c^\alpha
\quad
\forall c > 0,
\]
then we say $f$ is regularly varying with index $\alpha$.
$f$ can be written in the form $f(x) = {L(x) \over x^\alpha}$, where
$L(x)$ is a slowly varying function, i.e.
$\lim_{x \to \infty} {L(c x) \over L(x)} = 1$, $\forall c > 0$.
We call a random variable $X$ regularly varying with index
$\alpha \geq 0$ if it satisfies the tail balance condition in the
limit $x \to \infty$.
\[
\P(X > x) \sim p_+ {L(x) \over x^\alpha},\quad
\P(X < -x) \sim p_- {L(x) \over x^\alpha}, \quad
\text{for some } p_+, p_- \geq 0,\; p_+ + p_- = 1
\]

% Distribution functions, say $F(\cdot)$, whose survival function
% $\bar F(x) = 1 - F(x)$ satisfies the above scaling property, are
% observed in a variety of phenomena, 
When expanded to multiple dimensions, the scaling property of regular
variation is better described in terms of vague convergence to a
spectral measure $\mu_\alpha$: if a random vector $X$ satisfies
\[
{
  \P(X/|X| \in \cdot, |X| > c x)
  \over
  \P(|X| > x)
}
\overset{v}{\to} c^{-\alpha} \mu_\alpha(\cdot),
\quad
x \to \infty, \forall c > 0,
\]
then we say $X$ is regularly varying with index $\alpha$. Here
$\mu_\alpha$ is a probability measure on the unit sphere
\cite{buraczewski:damek:mikosch:2016}. It is called the spectral 
measure of $X$ and $\alpha$ is again the tail index.
If $X$ is regularly varying with index $\alpha$, then each component
and each linear combination of its components are regularly
varying with the same index $\alpha$. This follows from Feller
\cite{feller}, p. 278. Cf. also Jessen and Mikosch
\cite{jessen:mikosch:2006}, lemma 3.1, and Embrechts et. al.
\cite{embrechts:klueppelberg:mikosch:1997}, lemma 1.3.1.

@@

Random variables with regularly varying tails have some very nice
features: if $X_1$ and $X_2$ are both positive and regularly varying
with indices $\alpha_1$ and $\alpha_2$, respectively, then $a_1 X_1 + a_2 X_2$,
for $a_1, a_2 > 0$, is regularly varying with index
$\min\{\alpha_1, \alpha_2\}$ (cf. Mikosch and Jessen
\cite{jessen:mikosch:2006}). Moreover, if $X_1, X_2$ are iid,
$\P(a_1 X_1 + a_2 X_2 > u) \sim \P(a_1 X_1 > u) + \P(a_2 X_2 > u)$.

@@

In addition to Breiman's result, the following is also well-known,
assuming $X_1, X_2$ are positive independent random variables and
$X_1$ is regularly varying with index $\alpha$:
\begin{enumerate}
\item if $X_2$ is regularly varying with index $\alpha$ or
  $\lim_{x \to \infty} {P(X_2 > x) \over \P(X_1 > x)} = 0$, then $X_1
  X_2$ is regularly varying with index $\alpha$.
\item if $X_1, X_2$ are iid and $\E X_1^\alpha = \infty$, then
  ${\P(X_1 X_2 > x) \over \P(X_1 > x)} \to \infty$
\item if $X_1, X_2$ are iid and $\E X_1^\alpha < \infty$, then the
  only possible limit of ${\P(X_1 X_2 > x) \over \P(X_1 > x)}$ is
  $2 \E X_1^\alpha$.
\end{enumerate}
For an extensive summary of the regular variation properties of
functions of regularly varying random variables, see Mikosch and
Jessen \cite{jessen:mikosch:2006}.

@@

\section{Stochastic recurrence equation}
\label{sec:tyhyjyt}
One of the most important dynamical mechanisms that lead to regularly
varying random vectors is a stochastic recursion of the following form:
\begin{equation}
  \label{eq:rhjyu}
  X_t = A_t X_{t-1} + B_t, \quad t \in \integers
\end{equation}
where $X_t$ is a $d$-dimensional random vector, $A_t$ is a $d\times d$
random matrix and $B_t$ is a $d$-dimensional vector, random or
deterministic. The sequence $\{(A_t, B_t)\}_{t \in \integers}$ is
iid. The stationary solution to \eqref{eq:rhjyu} satisfies the fixed-point
equation $X \overset{d}{=} A X + B$, where $X$ and $(A, B)$ are
generic elements of the $\{X_t\}_{t \in \integers}$ and
$\{(A_t, B_t)\}_{t \in \integers}$ sequences.

@@

For discrete time models, Taylor \cite{taylor:1982} was the first to
propose a product process for modeling the martingale part $M_t$ of
$Y_t$:
\[
M_t - M_{t-1}= \sigma_t Z_t, \quad t \in \integers
\]
A convenient choice of $\sigma_t$ is
\begin{equation}
  \label{eq:rfht}
  \log\sigma_t = \sum_{l \in \integers} \psi_l \eta_{t-l},
  \quad
  t \in \integers
\end{equation}
where $\{\psi_l\}_{l \in \integers}$ is a sequence of real numbers
with at least one non-zero element,
$\{\eta_t\}_{t \in \integers}$ is an iid sequence of random
variables with zero-mean and finite variance.
By Kolmogorov's 3-series theorem, the infinite series above
converges if and only if $\sum_{l \in \integers} \psi_l^2 < \infty$,
and in this case $\{\sigma_{t}\}$ is stationary.
In particular, if $\{\eta_i\}_{i \in \integers}$ is normally
distributed with zero mean, $\{\sigma_t\}_{t \in \integers}$ has
log-normal marginal distributions.

@@

There are a few advantages in using stochastic volatility models.
They are among the simplest models allowing for conditional
heteroscedasticity (cf. Andersen et al.
\cite{andersen:davis:kreiss:mikosch:2009}); nevertheless,
they greatly improve the accuracy of predicting future volatilities
by taking advantage of high frequency data. Specifically, for the
semi-martingale $\{Y_t\}$ we have
\begin{equation}
  \label{eq:jieg}
  [Y_{\delta}]_t
  =
  \sum_{j=1}^{\floor{t/\delta}}
  (Y_{j \delta} - Y_{(j-1)\delta})^2
  \overset{P}{\to} [Y]_t,
  \quad
  \delta \to 0.
\end{equation}
The object $[Y_\delta]_t$ is called the realized quadratic variation
in time series literature.
The convergence in probability follows directly from the definition of
quadratic variation $[Y]_t$:
\[
[Y]_t = \lim_{\Delta t \to 0} \sum_{i=1}^{n} (Y_{t_i} - Y_{t_{i-1}})^2
\]
where $0 = t_0 < t_1 < t_2 < \cdots < t_n = t$ and
$\Delta t = \sup_{1\leq i \leq n} (t_i - t_{i-1})$.
%% When $\E Z_t = 0$, $[Y]_t = [M]_t$ 
Furthermore, Jacod \cite{jacod:1994} and
Barndorff-Nielsen and Shephard \cite{barndorff:shephard:2002} proved a
central limit theorem:
\[
  {[Y_\delta]_t - [Y]_t \over \sqrt{2 \delta \int_0^t \sigma_s^2 ds}}
  \overset{d}{\to} N(0, 1),
  \quad \delta \to 0.
\]
If the process of expected return has continuous sample path and $M_t$
is a stochastic volatility process, it has been shown that
$[Y]_t = [M]_t$ and
$[Y_\delta]_t \overset{P}{\to} \int_0^t \sigma_s^2 ds$.
Meanwhile, It\^o's formula gives, for a semi-martingale $\{Y_t\}$,
\begin{eqnarray}
  Y_t^2 &=& [Y]_t + 2 \int_0^t Y_s d Y_s \nonumber \\
  &=& [Y]_t + 2 \int_0^t Y_s dA_s + 2 \int_0^t Y_s dM_s, \nonumber \\
  \E Y_t^2 &=& \E [Y]_t + 2 \E \left( \int_0^t Y_s dA_s \right) \nonumber\\
  &\approx& \E [Y]_t, \quad \text{when $t$ is small}. \label{eq:hytre}
\end{eqnarray}
\eqref{eq:hytre} and \eqref{eq:jieg} show that $\E Y_t^2$, or in other
words, the forecast of future squared return, can be obtained as
$\E [Y_\delta]_t$, i.e. the forecast of future realized quadratic
variation.
% Since past realized quadratic variations are observable,
% forcast of future $[Y_\delta]_t$ can be obtained using e.g. an ARFIMA
% model.

@@

Conditional on $(Y_{t-1}, l_{t-1})$, the only source of randomness to
$Y_t$ comes from $Z_{t-1}$. Hence the shift of conditional probability
distribution shown in \eqref{eq:pgrtgpd} is equivalent to shifting the
distribution of $Z_{t-1}$. We have
\[
\P^\alpha (Z_{t-1} \in dz | Y_{t-1} = w) = |A(z^2) w|^\alpha {
  r_\alpha(A(z^2) \cdot w) \over r_\alpha(w)
} \P(Z_{t-1} \in dz).
\]
Note that $\{Z_t\}$ is an iid sequence in the original measure.
An expected value with respect to $\P^\alpha$ is related to its
counterpart with respect to $\P$ via
\begin{equation}
  \label{eq:pkjrkpo}
  \E [g(Y_t, l_t)] = \E^\alpha \left[
    g(Y_t, l_t) e^{-\alpha l_t} {r_\alpha(Y_{t-1}) \over r_\alpha(Y_t)}
    \right].
\end{equation}
Let us define $T_x = \min\{t \geq 1: |V_t| > x\}$. 
Once the first excursion of $|V_t|$ above $x$ has occurred, i.e.
$t > T_x$, we change the transition kernel back to the original and
continue the simulation until the process returns to a designated set
$\mathcal C = \{v: |v| \leq M\}$, where
the positive number $M$ is chosen in accordance with the function
$\Lambda$. We denote the successive times of $\{V_t\}$ returning to the
set $\mathcal C$ as $0 = K_0  < K_1 < K_2 < \cdots$. It can be shown that
$\{(V_{K_{m+1}}, \sum_{i=K_m}^{K_{m+1} - 1} \1{|V_i| > x})\}_{m \geq 0}$
is a positive Harris recurrent Markov chain for all $x \geq 0$.
Let $N_x = \sum_{i=0}^{K_1 - 1} \1{|V_i| > x}$. We show by the law of large
numbers for Markov chains
\[
\P(|V| > x) = \pi(\mathcal C) \E_\eta(N_x)
\]
where $\eta$ is $\pi$ restricted to the set $\mathcal C$, i.e.
$\forall S \subseteq \mathcal C, \eta(S) = \pi(S)/\pi(\mathcal C)$.
$\E_\eta$ means the expectation is taken only on condition
$V_0 \sim \eta$. Finally by \eqref{eq:pkjrkpo},
\[
\P(|V| > x) = \pi(\mathcal C) \E_\eta N_x
= \E_\eta^{\mathcal D} \mathcal E_x,
\]
where
\[
\mathcal E_x
=
\pi(\mathcal C)
N_x |A_{T_x} \cdots A_1 V_0|^{-\alpha}
{r_\alpha(Y_0) \over r_\alpha(Y_{T_x})}
\1{T_x < K_1}
\]
is our importance sampling estimator. $\E_\eta^{\mathcal D}$,
$\mathcal D$ for ``dual'', is to remind us that the expectation is
taken with respect to the shifted transition kernel as given by
\eqref{eq:pkjrkpo}, and then with respect to the original transition
kernel. In \S\ref{sec:efficiency} we show
\[
\limsup_{x \to \infty} {
  \E_\eta^{\mathcal D} \mathcal E_x^2
  \over
  [\P(|V| > x)]^2
} < \infty
\]
In plain words, the estimator $\mathcal E_x$ has bounded relative
error. See Asmussen and Glynn \cite{opac-b1123521}, \S 1 for
definition of {\em bounded relative error}.
The method presented in this paper is a multivariate generalization of
Collamore et al. \cite{collamore2014}. The reader is referred to it
for the treatment of the one-dimensional problem.
%% We propose an efficient importance sampling estimator for the rare
%%   event probability $\P(|V| > u)$ where $V$ is a vector following the
%%   stationary distribution of a GARCH($p, q$) process. Recall
%%   $V_t = A_t V_{t-1} + B$ is the matrix recurrence equation of a
%%   GARCH($p, q$) process. We emanate the process $\{V_t\}$ from a set
%%   $\mathcal C = \{V: |V| \leq M\}$ for some predefined positive
%%   constant $M$, and then we introduce a dual change of measure for the
%%   originally iid matrices $\{A_t\}$: for
%%   $t < T_u = \min\{i > 0: |V_i| > u\}$, we exponentially tilt the
%%   distribution of $A_t | X_{t-1}$, where $X_0 = V_0$ and
%%   $X_t = \prod_{i=1}^t X_0/|\prod_{i=1}^t X_0|$, so that
%%   $|A_t X_{t-1}|^\xi$ is more likely to take on large values and hence
%%   $|V_t|$ is more likely to exceed $u$. Once the exeedance has
%%   happened, we change the distribution of $\{A_t\}$ back to the
%%   original and continue the process until $V_t$ returns to the set
%%   $\mathcal C$. Along each simulated path emenating from $\mathcal C$
%%   and ending in $\mathcal C$ we compute $N_u = \sum_{i=1}^K \1{|V_i| >
%%     u}$, where $K = \min\{i > 0: |V_i| \leq M\}$.

@@

\subsection{Eigenvalues of the sample covariance matrix of a   
  stochastic volatility model}
In \S\ref{sec:SV:intro} we have introduced stochastic volatility
models and discussed their connection to realized quadratic
variation and their improved predictive power derived thereby. But
there we have only discussed univariate models. In fact, the
generalization to multivariate models is rather straightforward. In
chapter \ref{ch:bernoulli} we adopt the following multivariate model:
\begin{eqnarray*}
  X_{i,t} &=& \sigma_{i,t} Z_{i,t} \quad
  1 \leq i \leq p,\;
  t \in \integers, \\
  \log \sigma_{i, t}
  &=&
  \sum_{k,l \in \integers} \psi_{k,l} \eta_{i-k, t-l}\quad
  1 \leq i \leq p,\;
  t \in \integers,
\end{eqnarray*}
where $\{Z_{i,t}\}$ and $\{\eta_{i,t}\}$ are iid fields of random
variables. They are independent of each other. The distribution of
$\eta$, a generic element of $\{\eta_{i,t}\}$, satisfies
$\P(e^\eta > x) \sim x^{-\alpha} L(x)$,
where $\alpha > 0$ and $L(x)$ is a slowly varying function. The
coefficients $\{\psi_{k,l}\}$ are real and satisfy
$\sum_{k, l \in \integers} |\psi_{k,l}| < \infty$.

Depending on the tails of $\{\sigma_{i,t}\}$ and $\{Z_{i,t}\}$,
two situations can arise. When $\{Z_{i,t}\}_{t \in \integers}$
is a regularly varying sequence with index $\alpha \in (0, 4)$
and dominates the tail, we show that each of the sequences
$\{X_{i,t}\}_{t \in \integers}, 1 \leq i \leq p$ and each of
$\{X_{i,t} X_{j,t}\}_{t \in \integers}, 1\leq i < j \leq p$
are regularly varying with index $\alpha$, assuming suitable
conditions on $\{\sigma_{i,t}\}$.

@@

We are interested in the model
\begin{eqnarray}
  X_{i,t} &=& \sum_{k,l \in \integers} h_{k,l} Z_{i-k, t-l},
  \quad i,t \in \integers \label{eq:tyhjuj}
\end{eqnarray}
where $\{Z_{i,t}\}_{i,t \in \integers}$ is a field of iid random
variables and $h_{k,l}$ is an array of real coefficients. We
assume $\{Z_{i,t}\}$ is regularly varying with index
$\alpha \in (0, 4)$ and
\[
\sum_{k,l \in \integers} |h_{k,l}|^\delta < \infty
\]
for some $\delta \in (0, \min\{\alpha/2, 1\})$. This condition ensures
that the infinite series in \eqref{eq:tyhjuj} is almost surely
absolutely convergent. Since each $\{X_{i,t}\}_{t \in \integers}$
is a linear combination of the sequences $\{Z_{i,t}\}_{t \in \integers}$
that are regularly varying with index $\alpha$, each of
$\{X_{i,t}\}_{t \in \integers}$ is regularly varying with index $\alpha$.

@@

Corresponding to the time-lagged matrices $\mathbf X(s)$, we also introduce
the time-lagged coefficient matrices. Define
\[
\mathbf H(s) = \{h_{k, l+s}\}_{k, l \in \integers}
\]
and denote by $v_1(s) \geq v_2(s) \geq \cdots$ the singular values of
the matrix
$\mathbf H(0) \mathbf H(s) = \{\sum_{l \in \integers} h_{i,l} h_{j,l+s}\}_{i,j \in \integers}$.
We have established the following point process convergence result
upon appropriate conditions (see theorem \ref{cor:1} of chapter
\ref{ch:extremes}):
\[
\sum_{i=1}^p \varepsilon_{a_{np}^{-2}(\lambda_{(i)}(0), \ldots, \lambda_{(i)}(s))}
\overset{d}{\to}
\sum_{i=1}^\infty \sum_{j=1}^\infty
\varepsilon_{\Gamma_i^{-2/\alpha}(v_j(0), \ldots, v_j(s))}.
\]
where $\{a_{np}\}_{n,p \in \integers}$ is a sequence of positive real
numbers such that $n p \P(Z > a_{np}) \to 1, n \to \infty$.
$\varepsilon_{x}$ is the Dirac measure with unit mass at $x$ and
$\Gamma_i = \sum_{k=1}^i E_k$, where $\{E_k\}_{1 \leq k \leq i}$ is
an iid sequence of Exp($1$) random variables. In particular, we see
$a_{np}^{-2}\lambda_{(1)}(0) \overset{d}{\to} E_1^{-2/\alpha} v_1(0)$, i.e.
asymptotically $a_{np}^{-2}\lambda_{(1)}(0)$ has a scaled
Fréchet($\alpha/2$) distribution.
